Demography and linked selection in wild and domesticated maize

### 2015-10-19 manuscript.tex
typos and little things.

\begin{abstract}
  The genetic variability present in a population is increases through mutation and drift, and generally reduced via selection. But these forces do not act in a vacuum, and instead interact continuously to shape patterns of genetic diversity. To quantify the interplay between demography (which drives drift) and selection, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. Our sample included whole genome sequence data from 23 maize and 13 teosinte individuals.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps on genic mutations are not the primary force driving maize evolution. We find that a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels exceeding those seen in teosinte. Our results demonstrate that demographic and selective history in the ancient and recent past both contribute to the levels of genetic variability that are present today. Our results also show that for bottlenecked species, rapid post-bottleneck expansion can have wide-ranging impacts on diversity that conflict with the expected loss of diversity that occurs during the bottleneck itself.

@@

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to genetic changes \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. Importantly, demography and selection do not operate independently to determine genetic variability \cite{li2012}. Instead, these forces are engaged in a continuous battle by which each masks and obfuscates the outcomes of the other, ultimately meaning that the selective history of a species cannot be properly evaluated without a thorough estimate of demographic history, and demography cannot be evaluated without an understanding of selection.

@@

To that end, the objectives of our study were to 1) investigate the
relative importance of different forms of selection on whole-genome
variability, using maize and teosinte as a model; 2) research the impact that the
demographic history has had on genetic variability in maize, and how
this compares to the impact of a different demographic history in teosinte; and 3) precisely estimate the parameters of the
maize domestication bottleneck. Gaining an improved understanding of the interaction between demography and selection in maize will help to increase the understanding of these phenomenon across a wide array of organisms.

@@

\subsection{Hard sweeps do not explain diversity differences} %JRI
When selection increases the frequency of a new beneficial mutation, a signature of reduced diversity is left at surrounding linked sites \cite{smith1974}.
To evaluate whether patterns of such ``hard sweeps'' could explain observed differences in diversity between genic and intergenic regions of the genome, we compared diversity around missense and synonymous substitutions between \emph{Tripsacum} and either maize or teosinte.
If a proportion of missense mutations have been fixed due to hard sweeps, diversity around these substitutions should be lower than around synonymous substitutions. 
We observe this pattern around the causative amino acid substitution in the the domestication locus \emph{tga1} (Figure \ref{sFig:tga1}), likely the result of a hard sweep during domestication \cite{wang2005origin, wang2015}. Genome-wide, however, we observe no differences in diversity between synonymous and missense substitutions in either maize or teosinte (Figure \ref{fig:hardSweeps}).

@@

Selection can also reduce diversity in functional regions of the genome via removal of deleterious mutations, a process known as purifying or background selection \cite{charlesworth1993}.
We investigated purifying selection in maize and teosinte by evaluating the reduction of diversity around genes.
Pairwise diversity is strongly reduced within genes for both maize and teosinte (Figure \ref{fig:purify}A) but recovers quickly at sites outside of genes, consistent with the low levels of linkage disequilibrium generally observed in these subspecies \cite{tenaillon2002,chia2012}. 
The reduction in relative diversity is more pronounced in teosinte, reaching lower levels in genes and occurring over a wider region.

@@

\subsection{Population expansion leads to stronger purifying selection in modern maize} %JRI
Motivated by the rapid post-domestication expansion of maize evident in our demographic analysis, we reasoned that low-frequency --- and thus younger --- polymorphisms might show patterns distinct from pairwise diversity. 
Singleton diversity around missense and synonymous substitutions (Figure \ref{sFig:singleton}) appears nearly identical to results from pairwise diversity (Figure \ref{fig:hardSweeps}), providing little support for a substantial increase in the number or strength of hard sweeps occurring in maize. 
In contrast, we observe a significant shift in the effects of purifying selection: singleton polymorphisms are more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. This result is the opposite of the pattern observed for $\pi$, where teosinte demonstrated a stronger reduction of diversity in and around genes than did maize. Again, this relationship remained after removing the 20\% of genes with the highest H12 values (Figure \ref{sFig:H12}).
Finally, while direct comparison of pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, as indicated by our coalescent-based demographic analysis, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}).

@@

The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}. \cite{corbett2015} recently showed that selection tends to have an elevated impact reducing diversity in species with large population size compared to small, which may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently.

@@

A consequence of the previously-inefficient purifying selection in maize is that it likely harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. Therefore, high genetic load could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. This observation is not restricted to maize. Bottlenecks are have been documented or predicted across most domesticated plant and animal species \cite{doebley2006, hyten2006, zhu2007, bovine2009, larson2013 }, but rapid expansion post-bottleneck, which we demonstrate generates ubiquitous consequences relating to genetic diversity, is not well studied. In human, model, and other species, these results demonstrate that observed patterns of selection depend heavily on recent and/or ancient patterns of demography.

>>>>

\begin{abstract}
  The genetic variability present in a population results from a constant interaction between drift and selection. To quantify the interplay between demography (which drives drift) and selection, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. Our sample included whole genome sequence data from 23 maize and 13 teosinte individuals.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps on genic mutations are not the primary force driving maize evolution. We find that a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels exceeding those seen in teosinte. Our results demonstrate that demographic and selective history in the ancient and recent past both contribute to the levels of genetic variability that are present today. Our results also show that for bottlenecked species, rapid post-bottleneck expansion can have wide-ranging impacts on diversity that conflict with the expected loss of diversity that occurs during the bottleneck itself.

@@

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to changes in genetic variabiligy \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. Importantly, demography and selection do not operate independently to determine genetic variability \cite{li2012}. Instead, each of these forces is constantly masking and obfuscating the other, ultimately meaning that the selective history of a species cannot be properly evaluated without a thorough estimate of demographic history, and demography cannot be evaluated without an understanding of selection.

@@

To that end, the objectives of our study were to 1) investigate the
relative importance of different forms of selection on whole-genome
variability, using maize and teosinte as a model; 2) research the impact that
demographic history has had on genetic variability in maize, and how
this compares to the impact of a different demographic history in teosinte; and 3) precisely estimate the parameters of the
maize domestication bottleneck. Gaining an improved understanding of the interaction between demography and selection in maize will help to increase the understanding of these phenomena across a wide array of organisms.

@@

\subsection{Hard sweeps do not explain diversity differences} %JRI
When selection increases the frequency of a new beneficial mutation, a signature of reduced diversity is left at surrounding linked sites \cite{smith1974}.
To evaluate whether patterns of such ``hard sweeps'' could explain observed differences in diversity between genic and intergenic regions of the genome, we compared diversity around missense and synonymous substitutions between \emph{Tripsacum} and either maize or teosinte.
If a proportion of missense mutations have been fixed due to hard sweeps, diversity around these substitutions should be lower than around synonymous substitutions. 
We observe this pattern around the causative amino acid substitution in the the maize domestication locus \emph{tga1} (Figure \ref{sFig:tga1}), likely the result of a hard sweep during domestication \cite{wang2005origin, wang2015}. Genome-wide, however, we observe no differences in diversity between synonymous and missense substitutions in either maize or teosinte (Figure \ref{fig:hardSweeps}).

@@

In the case of purifying or background selection, diversity can be reduced in functional regions of the genome via removal of deleterious mutations \cite{charlesworth1993}.
We investigated purifying selection in maize and teosinte by evaluating the reduction of diversity around genes.
Pairwise diversity is strongly reduced within genes for both maize and teosinte (Figure \ref{fig:purify}A) but recovers quickly at sites outside of genes, consistent with the low levels of linkage disequilibrium generally observed in these subspecies \cite{tenaillon2002,chia2012}. 
The reduction in relative diversity is more pronounced in teosinte, reaching lower levels in genes and occurring over a wider region.

@@

\subsection{Population expansion leads to stronger purifying selection in modern maize} %JRI
Motivated by the rapid post-domestication expansion of maize evident in our demographic analysis, we reasoned that low-frequency --- and thus younger --- polymorphisms might show patterns distinct from pairwise diversity. 
Singleton diversity around missense and synonymous substitutions (Figure \ref{sFig:singleton}) appears nearly identical to results from pairwise diversity (Figure \ref{fig:hardSweeps}), providing little support for a substantial increase in the number or strength of hard sweeps occurring in maize. 
In contrast, we observe a significant shift in the effects of purifying selection: singleton polymorphisms are more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. This result is the opposite of the pattern observed for $\pi$, where teosinte demonstrated a stronger reduction of diversity in and around genes than did maize. As was the case when we analyzed $\pi$, this relationship remained after we removed the 20\% of genes with the highest H12 values (Figure \ref{sFig:H12}).
Finally, while direct comparison of pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, as indicated by our coalescent-based demographic analysis, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}).

@@

The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}. \cite{corbett2015} recently showed that selection tends to have an elevated impact reducing diversity in species with large population size compared to small, which may explain why maize and humans, both bottlenecked species, show similar patterns while \emph{drosophila} and \emph{capsella} behave differently.

@@

A consequence of the previously-inefficient purifying selection in maize is that it likely harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. Therefore, high genetic load could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. This observation is not restricted to maize. Bottlenecks have been documented or predicted across most domesticated plant and animal species \cite{doebley2006, hyten2006, zhu2007, bovine2009, larson2013 }, but rapid expansion post-bottleneck, which we demonstrate generates ubiquitous consequences relating to genetic diversity, is not well studied. In human, model, and other species, these results demonstrate that observed patterns of selection depend heavily on recent and/or ancient patterns of demography.
